Predicting Sequence Variability from Voronoi Tessellation of Proteins

### 2015-02-24 manuscript/voronoi/manuscript_voronoi.tex
minor changes to mn

    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural properties have been proposed over the past decade to answer this question. The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition. A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.  Expanding on these studies and using a dataset of 209 monomeric enzymes, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution. We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometrical average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates. We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.

@@

    Over the past decade, extensive attempt has been made to predict protein sequence evolution from structural properties.  A variety of site-specific structural characteristics have been proposed, including the Relative Solvent Accessibility (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins (xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014). \\

    Although structural characteristics have been individually extensively studied and explored, it is yet unknown whether these seemingly independent quantities are merely different manifestations of an underlying more fundamental characteristics of individual sites in proteins. It is perceivable that quantities such as B factor, RSA, and CN can all serve as {\it approximate} measures of local packing density of individual  sites in proteins, or the local flexibility of individual amino acids. Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the primary determinant of sequence variability, in contrast to local flexibility. Nevertheless, site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein. A more accurate measure of amino acid flexibility can be obtained from the accessible volume to each site in protein structure. An estimate of the accessible volume to each site can be obtained through Contact Number. The definition of this CN however, depends on an arbitrary parameter: the radius of neighborhood definition around each site.  An alternative more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure. \\

>>>>

    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural properties have been proposed over the past decade to answer this question. The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition. A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.  Expanding on these studies and using a dataset of 209 monomeric enzymes, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution. We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates.
    % We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.

@@

    Over the past decade, extensive attempt has been made to predict protein sequence evolution from structural properties.  A variety of site-specific structural characteristics have been proposed, including the Relative Solvent Accessibility (xx), Contact Number (xx), measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g., the quantity ddG rate in][]{}(xx echave 2014), and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) (xx) or flexibility measures based elastic network models (xx Bahar et al.) and Molecular Dynamics (MD) simulations (xx shahmoradi 2014). \\

    Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently. It is perceivable that quantities such as B factor, RSA, and CN can all serve as {\it approximate} measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids. Recently, Huang et al 2014 (xx) argued, through an extensive mathematical model, for Contact Number as the primary determinant of sequence variability, as opposed to local flexibility. Nevertheless, site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily a bais-free measure of the amino acid flexibility as a whole in a given site in protein. A more accurate measure of amino acid flexibility can be obtained from the accessible volume to each site in protein structure. An estimate of the accessible volume can be obtained through a quantity widely known as Contact Number (xx). For a given $C_\alpha$ backbone atom representing an amino acid in protein, the contact number is defined as the number of $C_\alpha$ atoms within a radius ($r$) of the reference $C_\alpha$ atom.  This definition of CN however, depends on an arbitrary parameter: the radius of neighborhood around the reference site.  An alternative more accurate definition of site-specific flexibility can be obtained from partitioning of the protein 3D structure. \\
